---
title: "Date and Time Classes in R: "
output: html_notebook
---

#Being a Short Training Material for Offa R Users Group Meeting on 27th March, 2025 at Statistical Laboratory, Statistics Department, The Federal Polytechnic Offa, Kwara State, Nigeria

#Prepared and Presented
#By
Udokang, Anietie Edem (Oganizer, ORUG)
Chif Lecturer, Statistics Department, The Federal Polytechnic Offa, Kwara State, Nigeria
##Introduction 
The focus of this short training shall be on how to handle date and time in R Statistical Software that allow for numerical and mathematical calculations.
There are three basic date/time classes in R. They are Date class, POSIXct class and POSIXlt class. It it is important to note that dates are stored in R internally as the number of days from 1970-01-01 and times are stored as the number of seconds from 1970-01-02.

##Date Class
The function as.Date is used to convert a character object to date object. This function by default can easily convert character objects to date objects as far they are in the following %Y-%m-%d and %Y/%m/%d.

#Example 1.1
```{r}
x<-"2020-01-25"
#x is a character object in R and not date object
x
```
```{r}
class(x)
```
```{r}
#to convert x from character object to date object named calendar_date we use as.Date function
calendar_date<-as.Date(x)
calendar_date
```
```{r}
#to view the class of calendar_date
class(calendar_date)
```
#Example 1.2
```{r}
x<-"2020/01/25"
#x is a character object in R and not date object
class(x)
```
```{r}
#to convert x from character object to date object named calendar_date we use as.Date function
calendar_date<-as.Date(x)
calendar_date
```
```{r}
#to view the class of calendar_date
class(calendar_date)
```
#Example 1.3
```{r}
#when x the character object is not in a recognizable format then format argument must be used
x<-"25/01/2020"
x
```
class(x)
```{r}
#to convert x to date object calendar_date we use as.Date function and the format argument
calendar_date<-as.Date(x, format="%d/%m/%Y")
calendar_date
```
```{r}
#to view the class of calendar_date
class(calendar_date)	
```
#Example 1.4
```{r}
#when x the character object is not in a recognizable format then format argument must be used
x<- "01232022"
x
```
```{r}
class(x)
```
```{r}
#to convert x to date object calendar_date we use as.Date function and the format argument
calendar_date <- as.Date(strptime(x, format = "%m%d%Y"))
calendar_date
```
```{r}
#to view the class of calendar_date
class(calendar_date)	
```

###POSITct
This function as.POSTct converts date-time objects to the POSIXct class, which stores date-time values as the number of seconds since January 1, 1970 (the Unix epoch). It’s useful for efficient storage and computation.
##Eample 2.1
```{r}
datetime <- as.POSIXct("2024-07-29 03:17:16")
print(datetime)
```
##Date and Time:
#Example 2.2

```{r}
datetime <- as.POSIXct("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%Y-%m-%d %H:%M:%S")
print(formatted_datetime)  
```
##Custom Date and Time Format:
#Example 2.3
```{r}
datetime <- as.POSIXct("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%d-%b-%Y %I:%M %p")
print(formatted_datetime)  
```
###POSITlt
This function as.POSITlt converts date-time objects to the POSIXlt class, which stores date-time values as a list with components for seconds, minutes, hours, days, months, years, etc. It’s more detailed but less efficient for storage compared to POSIXct.
#Example 2.4
```{r}
datetime <- as.POSIXlt("2024-07-29 03:17:16")
print(datetime)
```
2.3.1. Date and Time:
Example 2.5

```{r}
datetime <- as.POSIXlt("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%Y-%m-%d %H:%M:%S")
print(formatted_datetime)  

```
##2.2.2. Custom Date and Time Format:
#Example 2.6
```{r}
datetime <- as.POSIXlt("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%d-%b-%Y %I:%M %p")
print(formatted_datetime)  
```

###Differences between POSITct and POSITltnts of the date-time.
POSITct is used for efficient storage and when you don’t need to manipulate individual components of the date-time. Whereas POSITlt is used when you need to access or manipulate specific components like the hour or minutes. 

THANK YOU FOR YOUR PARTICIPATION
