---
title: "R’s Base Data Structures"
output: html_notebook
---
##Training Material for Offa-R-Users-Group (ORUG) Event Held on Saturday January 31, 2026 by 2.00pm at Statistics Laboratory, Statistics Department, Federal Polytechnic Offa, Nigeria

By

#Adebiyi Kazeem


#1. Introduction
• R is designed for statistical computing and data analysis.
• All R objects are built using base data structures.
• Understanding these structures is essential for efficient modelling.

#2. Atomic Vectors
• One-dimensional and homogeneous.
• All elements must be of the same type.
• Foundation of most other data structures.

#Types of Atomic Vectors
Logical, Integer, Double (numeric), Character, Complex, Raw

#Atomic Vector Example

```{r}
x <- c(1, 2, 3)
y <- c("A", "B", "C")
x
y
```

#3. Lists
• One-dimensional but heterogeneous.
• Can store different types and lengths.
• Commonly used for complex outputs.

#List Example
```{r}
my_list <- list(
  name = "R Presentation",
  year = 2026,
  scores = c(70, 85, 90))
my_list 
```
#4. Matrices
• Two-dimensional atomic vectors.
• Homogeneous elements.
• Used in linear algebra operations.

#Matrix Example
```{r}
M <- matrix(1:6, nrow = 2, ncol = 3)
M
```
#5. Arrays
• Generalisation of matrices.
• Multi-dimensional.
• Homogeneous data type.

#Array Example
```{r}
A <- array(1:12, dim = c(2, 3, 2))
A
```

#6. Data Frames
• Two-dimensional tabular structure.
• Columns may have different data types.
• Most common structure for datasets in R.

#Data Frame Example
```{r}
df <- data.frame(
  id = 1:3,
  age = c(25, 30, 28),
  gender = c("M", "F", "M"))
df
```
#7. Factors
• Represent categorical variables.
• Stored as integer codes with labels.
• Important in regression and ANOVA.

#Factor Example
```{r}
gender <- factor(c("Male", "Female", "Male"))
gender
```
#8. Comparison of Base Data Structures
Vector – 1D – Homogeneous
List – 1D – Heterogeneous
Matrix – 2D – Homogeneous
Array – ≥2D – Homogeneous
Data Frame – 2D – Heterogeneous (by column)
Factor – 1D – Homogeneous

#9. Conclusion
• Base data structures underpin all R objects.
• Correct usage improves efficiency and clarity.
• Essential knowledge for advanced R programming.

#THANK YOU FOR PARTCIPATING


