---
title: "RStudio Essentials"
output: html_notebook
---

#Being a Short Training Material used in Hybrid event at Offa R Users Group (ORUG) Meeting on 30th June, 2025 at Statistical Laboratory, Statistics Department, The Federal Polytechnic Offa, Kwara State, Nigeria


#Prepared and Presented By 
#Sodiq Ridwan
#Statistician and Data Scientist

#INTRODUCTION TO RSTUDIO
RStudio is a powerful Integrated Development Environment (IDE) which is built specifically for use with R programming language. It is free, open-source and user-friendly. Although R itself can be used through its basic R GUI (Graphical User Interface) which is installed by default but RStudio enhances the experience of the users by providing a modern, organized, as well as more efficient environment for coding, data analysis and report writing.

#SOME REASONS FOR USING RSTUDIO INSTEAD OF R GUI

```{r}
library(knitr)
table<-data.frame(Feature=c("Script Editor","Project Management", "Plots","Data Viewer","Reporting","Debugging","Appearance and Usability"),R_GUI=c("Basic and limited","Not available","Open in new windows","Basics","Not supported","Mannual","Outdated and plain"),RStudio=c("Full featured script editor with syntax highlighting and code folding","Built in support for R Projects and reproducibility","Appears in a dedicated pane, easier to manage","Spreadsheet-style viewer with filters","Supports RMarkdown, knitting to PDF/HTML/Word","Built-in debugging tools and console tracking","Modern, customizable,and user-friendly interface"))
kable(table,caption = "SOME REASONS FOR USING RSTUDIO INSTEAD OF R GUI")
```
#RSTUDIO INTERFACE
1 – Source Editor: This is where you write, edit, and save your R script
2 – Environment: This displays variables in memory and previously executed commands
3 – Console: This is where code is executed and outputs are displayed
4 – Files/Plots/Packages/Help/Viewer/Presentation: This lets you navigate folders, view plots, manage packages, and access documentation. 
#UNDERSTANDING THE TOP MENU BAR IN RSTUDIO
The top menu bar in RStudio provides access to a variety of essential tools that help you write code, manage projects, run sessions, customize your view, and troubleshoot. Each menu item contains grouped functions designed to streamline specific tasks. 
#Here is the breakdown of some of the most useful ones:
#File Menu
The File menu is where you manage your scripts, projects, and documents. Here are some options you will find in File Menu and a brief description of their uses:
New File – Create a new R Script, RMarkdown, or Notebook.
Open File – Open and existing .R or .Rmd file.
Save/Save As – Save your current script.
New Project – Start a new RStudio project (very useful for organizing work).
#Edit Menu
The Edit menu helps you manipulate code in the script editor. Here are some options you will find in Edit Menu and a brief description of their uses:
Undo/Redo – Undo or redo your action.
Cut/Copy/Paste – Standard test manipulation.
Find/Replace - Search or replace words within a script.
 
#Session Menu
The Session menu is essential for managing your R session such as clearing memory, restarting R, or changing your working directory. Here are some options you will find in Session Menu and a brief description of their uses:
Restart R – Clears all memory and restarts R (great for resetting your session).
Terminate R – Stops the current session.
Set Working Directory – Choose or change the default file folder.
Clear Workspace – Removes all objects from memory.
#Help Menu
Provides built-in documentation and access to learning resources. Here are some options you will find in Help Menu and a brief description of their uses:
R Help – Opens the help viewer.
Keyboard Shortcuts Help – View or print all shortcut keys
Cheat Sheets – Download official Posit (RStudio) cheat sheets
About RStudio – Version and license information
#CREATING AND RUNNING FIRST SCRIPT IN RSTUDIO
Go to File → New File → R Script
In the new script window, type this code:
# My first R code in RStudio

```{r}
name <- "RStudio Essentials"
print(name)
```
Place your cursor on any line and press Ctrl + Enter (or Cmd + Enter on Mac).
Output: You will see the result ([1] "RStudio Essentials") in the Console pane.
#SAVE THE SCRIPT
Click File → Save As
Save as “demo_script.R” to your working folder
#CREATING A SIMPLE DATA FRAME

```{r}
students <- data.frame(
  Name = c("John", "Ada", "Ben"),
  Score = c(85, 92, 78),
  Age = c(20, 21, 22)
)
View(students)# Opens a spreadsheet-like viewer
kable(students) #to display the output below
```
#Observation:
The students data frame appears in the Environment tab.
View() opens it in a spreadsheet style in a new tab.
#PERFORM A SUMMARY AND PLOT
```{r}
summary(students)    # Summary statistics
hist(students$Score) # Histogram in Plots pane
```

Output:
A quick summary of numeric columns.
A histogram appears in the Plots pane.


