---
title: "WORKING WITH FACTORS AND CATEGORICAL DATA IN R"
output: html_notebook
---

#Being a Short Training Material used in online event at Offa R Users Group (ORUG) Meeting on 30th September, 2025 


#Prepared and Presented By 
#Sodiq Ridwan
#Statistician and Data Scientist

#OUTLINE:
1. WHAT IS CATEGORICAL DATA?
2. WHAT IS A FACTOR IN R?
3. HOW TO CREATE FACTORS, REORDER AND RENAME LEVELS
4. WHY UNDERSTANDING CATEGORICAL DATA AND FACTORS IS IMPORTANT?
5. IMPLICATIONS OF NOT UNDERSTANDING OR DECLARING FACTORS
6. AREAS WHERE FACTORS ARE USEFUL
7. KEY TAKEWAYS
#1. WHAT IS CATEGORICAL DATA?
Categorical data refers to data that represents groups, categories, or labels.
#Examples in real life:
a)	Gender: Male, Female
b)	Blood Type: A, B, AB, O
c)	Education Level: Primary, Secondary, Tertiary
d)	Grades: A, B, C, D
#Categorical data can be:
a) Nominal – categories with no natural order (e.g., Male/Female, Blood type)
b) Ordinal – categories with a natural order (e.g., Low, Medium, High or Grades A > B > C).

#2. WHAT IS A FACTOR IN R?
a)	A factor in R is a data structure used to represent categorical data.
```{r}
blood_group <- factor(c("A", "O", "B", "A","AB","O","A","AB"))
blood_group
```
Here:
The data points are (“A”, “O”, “B”, “A”,“AB”,“O”,“A”,“AB”))
The levels (categories) are “A”, “B”, “AB” and “O”.

#3. HOW TO CREATE FACTORS, REORDER AND RENAME LEVELS
#a) How to create factors
```{r}
size <- factor(c("medium", "small","small","small","large", "medium", "small"))
summary(size)
```
#b) How to check levels
```{r}
levels(size)
```
#c) How to Reorder Levels
```{r}
size <- factor(size,
               levels = c("small", "medium", "large"),
               ordered = TRUE)
levels(size)
```
```{r}
summary(size)
```
#d) Rename levels
```{r}
levels(size)<-c("size 38","size 42","size 46")
size
```
```{r}
summary(size)
```
#4. WHY UNDERSTANDING CATEGORICAL DATA AND FACTORS IS IMPORTANT?
#General importance:
a)	Many real-world datasets contain categorical variables (gender, region, product type).
b)	Statistical methods treat categories differently from numeric variables.
In R context:
a)	R uses factors in modeling (e.g., regression, ANOVA).
b)	Summaries and plots rely on correctly defined factors.
c)	Without factors, categorical data may be misinterpreted as characters (strings) or numeric codes, leading to wrong analysis results.

#5. IMPLICATIONS OF NOT UNDERSTANDING OR DECLARING FACTORS
#a) Misleading Summaries
```{r}
gender <- c("Male", "Female", "Female", "Male")
summary(gender)   # Only shows class and length
```
With factor:
```{r}
gender_factor <- factor(gender)
summary(gender_factor)
```
Without factors, you don’t see category counts.
#b) Wrong Ordering in Plots
```{r}
months <- c("Mar", "Jan", "Jan", "Mar", "Feb", "Jan")
barplot(table(months))
```
With factor (ordered):
```{r}
months_factor <- factor(months, levels = c("Jan", "Feb", "Mar"), ordered = TRUE)
barplot(table(months_factor))
```
Ensures logical order (A → B → C).
#c) Misinterpretation in Models
#Create dataset
```{r}
df <- data.frame(
  Group = c(1, 2, 1, 3, 2, 3, 1, 2, 3,2,3,2,3,1),
  Passed = c(1, 1, 0, 0, 1, 0, 1, 0, 1,0,0,0,0,1)
)
```
# Logistic regression treating Group as numeric
```{r}
model_wrong <- glm(Passed ~ Group, data = df, family = binomial)
summary(model_wrong)
```
# Convert Group to factor
```{r}
df$Group <- factor(df$Group, levels = c(1, 2, 3))
```
# Logistic regression with Group as categorical predictor
```{r}
model_correct <- glm(Passed ~ Group, data = df, family = binomial)
summary(model_correct)
```
Factors are essential in regression and ANOVA.
#6. AREAS WHERE FACTORS ARE USEFUL
#a) Summaries
```{r}
summary(factor(c("Yes","No","Yes","Yes")))
```
Gives counts of Yes and No.
b) Plots
```{r}
colors <- factor(c("Red","Blue","Red","Green","Blue"))
barplot(table(colors))
```
Bar chart grouped by category
#c) Statistical Models
```{r}
employees <- data.frame(
  Gender = factor(c("Male", "Female", "Male", "Female", "Male", "Female")),
  Salary = c(50000, 55000, 52000, 60000, 48000, 62000)
)

model_1<-lm(Salary ~ Gender, data = employees)
summary(model_1)
```
#7. KEY TAKEWAYS
a)	Categorical data = data divided into groups or labels.
b)	Factors in R = special data structure for categorical data.
c)	Without factors, R may give misleading summaries, wrong plots, and flawed models.
d)	Factors are essential in summaries, visualizations, and modeling.

#Always declare categorical variables as factors for accurate and meaningful analysis in R.

#Thank you for participating
#Hope to see you in the next meeting/training

#Visit: https://www.meetup.com/fedpofa-r-users-group/


